Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-18 manuscript/manuscript.tex
Working on adult results section

\textbf{Methods:} 6939 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/GPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC to physicians among models were evaluated for all records and by COD, age group, and age ranges.

\textbf{Results:} Based on PCCC, GPT-4 had the best performance overall (0.61), followed by GPT-3.5 (0.58), InSilicoVA (0.44), and InterVA-5 (0.43). GPT-4 also had the best performance for the adult (0.65), child (0.57), and neonatal (0.62) age groups, while performance decreased as adults aged (0.72-0.59) and increased as children and neonates aged (0.5-0.7). 7 of 19, 4 of 10, and 1 of 7 adult, child, and neonatal CODs respectively had models with $\geq$0.8 PCCC. For most CODs, GPT models had one of the highest performances with the exception of InSilicoVA for adult tuberculosis, child pneumonia and neonatal infections. At the population level, all models had CSMF accuracy between 0.7-0.79.

@@

Initially, 11,920 records from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \citep{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation.

@@

This section details the performance results of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CGHR-10 CODs after applying the methods in Section \ref{methods}. GPT-4 performed the best overall at 0.61 PCCC followed by GPT-3.5 at 0.58 PCCC. GPT-4 also had the highest PCCC for most age ranges and CODs across the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups with GPT-3.5, InterVA-5, and InSilicoVA having higher PCCC values for a few age ranges and CODs. Overall performance results are seen in Section \ref{results-overall}, and performance by adult, child, and neonatal records are seen in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo} respectively.

@@

Of the 6939 records, GPT-4 had the highest PCCC followed by GPT-3.5, InSilicoVA, and InterVA-5 at 0.61, 0.56, 0.44, and 0.43 PCCC respectively (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.03 PCCC. CSMF accuracies were above 0.7 for all models at 0.79, 0.74, 0.74, and 0.78 for GPT-4, GPT-3.5, InSilicoVA, and InterVA-5 respectively. Figure \ref{fig-pccc} shows the PCCC performance across age groups with reference to the CSMF Accuracy. GPT-4 had the highest PCCC at 0.65 and 0.62 for the adult and neonatal age groups, while GPT-3.5 had the highest PCCC at 0.57 for the child age group with GPT-4 performing slightly worse at 0.54. GPT-3.5 had the second highest PCCC at 0.56 fo the adult age group, while InSilicoVA had the second highest PCCC for the neonatal age group at 0.56. InSilicoVA and InterVA-5 performed the worse for both adult and child age groups at PCCC scores below 0.5, while GPT-3.5 performed the worse for the neonatal age group with a PCCC of 0.42. The PCCC for models had larger differences between the minimum and maximum PCCC for the adult age group at 0.24, than the child and neonatal age groups at 0.12 and 0.20 respectively. For reference, all records, including non-physician agreed records, were also coded by GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for comparison to physician agreed records. For physician agreed records, models had 0.05 to 0.09 higher PCCC values and 0.01 to 0.05 higher CSMF accuracy values than for all records.

@@

@@

Figure \ref{fig-pccc-cod-adult} shows the highest and lowest performing models by PCCC across 17 adult CODs excluding suicide due to the less notable 3 (\textless 1\%) records. The highest PCCC values were GPT-4 between 0.36 to 0.99 for 10 of 17 CODs (GPT-3.5, InterVA-5, and InSilicoVA within 0.05 for 3, 1, and 1 of these 10 CODs respectively), InSilicoVA for tuberculosis and road and transport injuries at 0.72 and 0.84 (GPT-4 and InterVA-5 within 0.05), and GPT-3.5 between 0.44 and 0.94 for 5 of 17 CODs (GPT-4 within 0.05 for malaria). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 8 of 17 CODs (InSilicoVA and GPT-3.5 within 0.05 for cancers and unspecified infections respectively), InSilicoVA between 0.01 to 0.41 for 6 of 17 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 at 0.39 for tuberculosis and 0.54 for road and transport injuries. Chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 0.93. Maternal conditions, and road and transport injuries, had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC below or at 0.2. Ischemic heart disease and unspecified infections had PCCC values below or at 0.65, where there were small differences between minimum and maximum PCCC at or below 0.14. Each adult COD had between 27 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, and other CODs between 1-9\% of all records.

@@

\subsubsection{Performant for Particular Causes of Death}\label{discuss-adv-cod}

>>>>

\textbf{Methods:} 6939 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for GPT-3.5/4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC to physicians among models were evaluated for all records and by COD, age group, and age ranges.

\textbf{Results:} Based on PCCC, GPT-4 had the best performance overall (0.61), followed by GPT-3.5 (0.58), InSilicoVA (0.44), and InterVA-5 (0.43). GPT-4 also had the best performance for the adult (0.65), child (0.57), and neonatal (0.62) age groups, while performance decreased as adults aged (0.72-0.59) and increased as children and neonates aged (0.5-0.7). 7 of 19, 4 of 10, and 1 of 7 adult, child, and neonatal CODs respectively had models with $\geq$0.8 PCCC. For most CODs, GPT-4 models had one of the highest performances with the exception of InSilicoVA for adult tuberculosis, child pneumonia and neonatal infections. At the population level, all models had CSMF accuracy between 0.7-0.79.

@@

Initially, 11,920 records from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \citep{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). For each record, two codes were assigned by two different randomly selected physicians, where codes were evaluated for agreement using Central Medical Evaluation Agreement 10 (CMEA-10) codes. CMEA-10 groups a range of similar ICD-10 codes together, where if they are in agreement if they are within the same group \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation.

@@

This section details the performance results of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CGHR-10 CODs after applying the methods in Section \ref{methods}. GPT-4 performed the best overall at 0.61 PCCC followed by GPT-3.5 at 0.56 PCCC. GPT-4 also had the highest PCCC for most age ranges and CODs across the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups with GPT-3.5, InterVA-5, and InSilicoVA having higher PCCC values for a few age ranges and CODs. Overall performance results are seen in Section \ref{results-overall}, and performance by adult, child, and neonatal records are seen in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo} respectively.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF).

Figure \ref{fig-pccc} shows the PCCC performance across age groups. GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC).

Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}. PCCC rose for records with ages from 0 to 27 days, dropped drastically for 1-5 months, rose again from 6 months to 14 years, and slowly declined for ages 15 to 69 years. The highest and lowest performances were observed for ages 10-29 years and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{fig-pccc-age.pdf}
\caption{Model performance by age range.}\label{fig-pccc-age}
\end{figure}

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide, due to a low 3 (\textless 1\%) records. GPT-4 had the highest individual performance for 10 of 17 CODs (0.36 to 0.99 PCCC), InSilicoVA for 2 CODs (0.72 and 0.84 PCCC), and GPT-3.5 for 5 CODs (0.44-0.94 PCCC). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 8 of 17 CODs (InSilicoVA and GPT-3.5 within 0.05 for cancers and unspecified infections respectively), InSilicoVA between 0.01 to 0.41 for 6 of 17 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 at 0.39 for tuberculosis and 0.54 for road and transport injuries. Chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 0.93. Maternal conditions, and road and transport injuries, had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC below or at 0.2. Ischemic heart disease and unspecified infections had PCCC values below or at 0.65, where there were small differences between minimum and maximum PCCC at or below 0.14. Each adult COD had between 27 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, and other CODs between 1-9\% of all records.

@@

\subsubsection{Performance for Particular Causes of Death}\label{discuss-adv-cod}
